---
title: "Capstone I"
author: "Francese"
date: "2023-12-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import & Data

```{r cars}
read_csv("23_WCSU_aims.csv")
aims_23 <- as.data.frame(read_csv("23_WCSU_aims.csv"), show_col_types = FALSE)
```

```{r}
aims_23<- aims_23[-c(1:3,7,9:10)]
aims_23
```

```{r}
colnames(aims_23)[1]<-"SchoolName"

aims_23$RiskStatus <- factor(aims_23$RiskStatus, levels = c("HIGH", "MODERATE", "LOW"))

aims_23$Period <-factor(aims_23$Period, levels = c('Fall','Winter','Spring'))
```

## Recreating Tables from Python

```{r}
#battery_scores<- aims_23 %>%
  #filter(IsBatteryScore == 1)
battery_scores
```

```{r}
spring_batteries <- battery_scores %>%
  filter(Period == 'Spring')

spring_batteries
```

```{r}
skill_scores <- aims_23 %>%
  filter(IsBatteryScore == 0)

skill_scores_grade_1 <- skill_scores %>%
  filter(StudentGrade == 1)
skill_scores_grade_1 

skill_scores_grade_1_bbe <- skill_scores_grade_1 %>%
  filter(SchoolName == "Blue Bird Elementary School")

unique(skill_scores_grade_1_bbe$MeasureName)
unique(skill_scores_grade_1_bbe$RelatedForm)
lit_skills_grade_1_bbe <- skill_scores_grade_1_bbe %>%
  filter(RelatedForm == "LWSF" | RelatedForm == "ORF" |RelatedForm == "WRF" |
           RelatedForm == "NWF"| RelatedForm == "PS"| RelatedForm ==  "AV" |
           RelatedForm == "LNF"|RelatedForm == "IS"  )
```

```{r}
high_risk <- battery_scores%>%
  group_by(RiskStatus)%>%
  summarise(
    num_students= length(unique(Students)))

high_risk
```

```{r}
students_at_risk<-battery_scores%>%
  group_by(MeasureName,RiskStatus)%>%
  summarise(
    num_students= length(unique(Students)), na.rm = TRUE)
students_at_risk$percent_of_group<-round(students_at_risk$num_students/414, digits=2)

arrange(students_at_risk, MeasureName, RiskStatus, desc(num_students))
```

```{r}
students_at_high_risk<-battery_scores%>%
  group_by(MeasureName,RiskStatus)%>%
  summarise(
    num_students= length(unique(Students)))%>%
  filter(RiskStatus=="HIGH")

students_at_high_risk$percent_of_group <- round(students_at_high_risk$num_students/414, digits=2)
students_at_high_risk
```

```{r}
high_risk_periods <-battery_scores%>%
  group_by(Period,RiskStatus)%>%
  summarise(
    num_students= length(unique(Students)))%>%
  filter(RiskStatus=="HIGH")

high_risk_periods
```

```{r}
tested_students_by_grade <- battery_scores%>%
  group_by(StudentGrade) %>%
  summarise(
    num_students= length(unique(Students)))

  
  tested_students_by_school <-
    battery_scores%>%
  group_by(SchoolName) %>%
  summarise(
    num_students= length(unique(Students)))
  
  
  tested_students_by_grade_school <-
    battery_scores%>%
  group_by(SchoolName, StudentGrade) %>%
  summarise(
    num_students= length(unique(Students)))
  tested_students_by_grade_school
```

```{r}
```

```{r}
grade_level_risk_spring <- battery_scores%>%
  group_by(StudentGrade, Period, MeasureName, RiskStatus) %>%
  summarise(
    num_students= length(unique(Students)),
.groups = "keep")%>%
  filter(Period=="Spring", RiskStatus =="HIGH")

grade_level_risk_spring
```

```{r}
first_grade_lit_scores <- battery_scores %>%
  group_by(Students) %>%
  filter(Period =="Spring", StudentGrade==1, MeasureName=="Early Literacy")

first_grade_lit_scores
```

## Merge Tables

```{r echo=TRUE}
spring_high_grade<- merge(grade_level_risk_spring, tested_students_by_grade, by = "StudentGrade")
#spring_high_grade$num_high_risk <- spring_high_grade$num_students.x 
#spring_high_grade$ttl_students_tested <- spring_high_grade$num_students.x 
#spring_high_grade$percent_of_pop <- round(num_high_risk/ttl_students_tested, digits = 2)


colnames(spring_high_grade)[5]<-"HighRiskStudents"
colnames(spring_high_grade)[6]<-"TotalTested"

spring_high_grade$percent_of_group <- round(spring_high_grade$HighRiskStudents/spring_high_grade$TotalTested, digits=2)
spring_high_grade 
```

## Plotting Results

```{r}
my_theme <- theme_minimal() + theme(text = element_text(face = "bold", family = "sans"))
dis_colors <- list(c("#9e3434","#fcbe03", "#54a157"))
#scale_fill_manual(values=c("#9e3434", "#54a157","#fcbe03"))
```

```{r}
sb<-spring_batteries  %>% 
  ggplot(mapping = aes(y=Natl, x=StudentGrade, fill = MeasureName)) +
  geom_boxplot() + scale_fill_brewer(palette = "BuBr")
sb <- sb + labs(title = "Standardized Student Battery Scores", y = "Natl Percentile", x = "Student Grade Level") + 
  theme (aspect.ratio = 1)  + my_theme

sb
```

b \<- b+ facet_grid(row=vars(StudentGrade))

```{r}
b<-spring_batteries  %>% 
  ggplot(mapping = aes(y=Natl, x=Dist, color= MeasureName),color="black") +
  geom_jitter() + scale_fill_brewer(palette = "Dark2")
b <- b + labs(title = "Standardized Student Battery Scores", y = "Natl Percentile", x = "Dist Percentile") + 
  theme (aspect.ratio = 1)  + my_theme+ 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "italic"))


b
```

```{r}

ggplot (students_at_risk) +
    geom_col(mapping = aes(x= MeasureName, fill= RiskStatus, y=num_students), color = "black") +
  theme (aspect.ratio = 1) + scale_fill_manual(values=c("#9e3434","#fcbe03", "#54a157")) +
  labs(x = NULL, y = NULL) + my_theme + labs(title = "Risk Levels of Students", x= "Assessment", y= "Students")
```

```{r}
high_risk_periods$Period <-factor(high_risk_periods$Period, levels = c('Fall','Winter','Spring'))

ggplot (high_risk_periods) +
    geom_col(mapping = aes(x= Period, y=num_students, fill=Period), color = "black") +
  theme (aspect.ratio = 1) + scale_fill_brewer(palette = "BrBG") +
  labs(x = NULL, y = NULL) + my_theme + labs(title = "Risk Levels of Students by Period", x= "Period", y= "Students")
```

scale_fill_brewer(palette = "BrBG")

```{r}
#grade_level_risk_spring$StudentGrade <-factor(grade_level_risk_spring,levels = c(1,2,3,4,5,6,7,8))
 glrs<-ggplot(grade_level_risk_spring) +
  geom_col(aes(x=StudentGrade, y=num_students, fill=MeasureName), color="black") +
  labs (x = "Assessment", y = "Count of High Risk Students", title = "Students Ranked 'High Risk' in Spring Time Assessments") + scale_fill_brewer(palette = "BrBG") +my_theme 
glrs
```

```{r}

ggplot(spring_high_grade, aes(x=StudentGrade, y=HighRiskStudents)) +
  geom_col(aes(fill=MeasureName), color="black") +
  labs (x = "Assessment", y = "Count of High Risk Students", title = "Students Ranked 'High Risk' in Spring Time Assessments") + scale_fill_brewer(palette = "BrBG") +my_theme + 
  geom_point()+
  geom_segment(aes(x=StudentGrade, xend=StudentGrade, y=0, yend=TotalTested), color="black")

```

```{r}
spring_high_grade 

geom_point(spring_high_grade)
```

```{r}
school_1_lit = first_grade_lit_scores %>%
  group_by(SchoolName, RiskStatus) %>%
  summarise(row_count=n())

school_1_lit$short_schoolname <- sapply(strsplit(school_1_lit$SchoolName, " "), `[`,1)
school_1_lit <- school_1_lit %>%
  group_by(short_schoolname) %>%
  mutate(total = sum(row_count)) %>%
  ungroup() %>%
  mutate(percent = (row_count / total) * 100)
school_1_lit
```

```{r}
ggplot(school_1_lit, aes(x=short_schoolname, y=row_count, fill= RiskStatus)) +
  geom_bar(stat="identity", color="black") + scale_fill_manual(values=c("#9e3434","#fcbe03", "#54a157")) + my_theme  + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "italic"))  +
  labs (x = "School Name (abbr)", y = "Count of Students", title = "First Grader Risk Level in Spring Literacy Assessments") 

```

```{r}

ggplot(school_1_lit, aes(x = short_schoolname, y = row_count, fill = RiskStatus)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("#9e3434", "#fcbe03", "#54a157")) +
  geom_text(aes(label = paste0(round(percent, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 3, fontface="bold") +
  my_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "italic")) +
  labs(x = "School Name (abbr)", y = "Count of Students", 
       title = "First Grader Risk Level in Spring Literacy Assessments")
```

```{r}
school_1_lit_bbe_2 <-school_1_lit %>%
  filter(SchoolName == "Blue Bird Elementary School")
school_1_lit_bbe_2 
```

```{r}

school_1_lit_bbe_2 <- battery_scores %>%
  filter(SchoolName == "Blue Bird Elementary School", StudentGrade==1, RelatedForm == 'ELB')
school_1_lit_bbe_2 %>%
  ggplot(aes(y=percent, x=row_count, fill=RiskStatus)) +
  geom_col(color="black") + scale_fill_manual(values=c("#9e3434","#fcbe03", "#54a157")) + my_theme  + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "italic"))   
```

```{r}
lit_skills_grade_1_bbe_spring
```

```{r}
#look at Blue Bird Elementary School skill scores
library(ggridges)

lit_skills_grade_1_bbe_spring<- lit_skills_grade_1_bbe %>%
  filter(Period== "Spring")

ggplot(lit_skills_grade_1_bbe_spring) +
  geom_density_ridges(aes(x = Natl, y = MeasureName, fill = MeasureName), scale = 0.6) +
  scale_fill_brewer(palette = "BrBG")  + my_theme +
  labs(x = "Percentile", y = "Skill Measure Name", 
       title = "Blue Bird Elementary Grade 1 Percentile Ranks") +
  geom_density_ridges(aes(x=Dist, y=MeasureName, fill= MeasureName), scale = 0.6, alpha = 0.5) +
  scale_fill_brewer(palette = "BrBG")  + my_theme


```

```{r}
lit_skills_grade_1_by_stu <- lit_skills_grade_1_bbe_spring%>%
  group_by(MeasureName) %>%
  summarise(num_students= length(unique(Students)), avg_score = mean(Score),
            avg_natl_per = mean(Natl), avg_dist_per = mean(Dist))
lit_skills_grade_1_by_stu
```

```{r}
lit_skills_grade_below_50 <- lit_skills_grade_1_bbe_spring%>%
  filter(Natl <= 50 & Dist <= 50)
lit_skills_grade_below_50
```

```{r}
skill_scores_avg <- skill_scores %>%
  group_by(Students, SchoolName, StudentGrade, Period, MeasureName, RelatedForm) %>%
  summarize (avg_score = mean(Score),
             row_count=n(),
             avg_natl = mean(Natl),
             avg_dist = mean (Dist)) %>%
  filter(Period == "Spring")

skill_scores_avg 
  
```

```{r}
first_grade_lit
```

```{r}
unique(skill_scores$RelatedForm)
```

```{r}
lit_skill_scores <- skill_scores%>%
  filter(RelatedForm == "LWSF" | RelatedForm == "ORF" |RelatedForm == "WRF" |
           RelatedForm == "NWF"| RelatedForm == "PS"| RelatedForm ==  "AV" |
           RelatedForm == "LNF"|RelatedForm == "IS"| RelatedForm == "RC" |
          RelatedForm == "SRF"| RelatedForm == "VOC"| RelatedForm == "PS" |
           RelatedForm == "IS")
```

```{r}
lit_skill_scores_avg = lit_skill_scores %>%
  group_by(StudentGrade, MeasureName, Period) %>%
  summarize (avg_score = mean(Score),
             num_students = length(unique(Students)),
             avg_natl = mean(Natl),
             avg_dist = mean (Dist),.groups="keep") %>%
  filter(Period == "Spring")
lit_skill_scores_avg
```

```{r}
unique(lit_skill_scores$SchoolName)

lit_skill_scores_bbms<- lit_skill_scores %>%
  filter(SchoolName == "Brown Bear Middle School")

lit_skill_scores_wes<- lit_skill_scores %>%
  filter(SchoolName == "Whipple Elementary School\n Elementary School")


lit_skill_scores_bbe<- lit_skill_scores %>%
  filter(SchoolName == "Blue Bird Elementary School")

lit_skill_scores_wre<- lit_skill_scores %>%
  filter(SchoolName == "White Rabbit Elementary School\n Elementary School")

lit_skill_scores_dde<- lit_skill_scores %>%
  filter(SchoolName == "Double Dog Elementary School")

lit_skill_scores_grp <- lit_skill_scores %>%
  group_by(SchoolName,StudentGrade, MeasureName, Period) %>%
  summarise(num_students = length(unique(Students)),
            avg_score = mean(Score),
            avg_natl=mean(Natl)
            )
lit_skill_scores_grp

cleaned_lit_skills <- na.omit(lit_skill_scores_grp)


cleaned_lit_groupedskills<- cleaned_lit_skills %>%
  group_by(SchoolName, StudentGrade, Period)%>%
  summarise(num_students = sum(num_students),
            avg_score = mean(avg_score),
            avg_natl = mean(avg_natl))
cleaned_lit_groupedskills
```

```{r}
cleaned_lit_groupedskills$Period <- factor(cleaned_lit_groupedskills$Period, 
                                           levels = c("Fall", "Winter", "Spring"))

```

```{r}

cleaned_lit_groupedskills %>%
  ggplot(aes(x = Period, y = avg_natl, color = SchoolName)) +
  geom_point() +
  geom_line() +
  my_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "italic")) +
  theme(text = element_text(size = 9), element_line(size =1)) +
  labs(y = "National Percentile", x = "Student Grade", title = "Spring Literacy Skills")

```

```{r}
cleaned_lit_groupedskills %>%
  ggplot(aes(y = Period, x = avg_natl, group = SchoolName, color = SchoolName)) +
  geom_point() +
  geom_line() +
  my_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "italic")) +
  theme(text = element_text(size = 9), element_line(size = 1)) +
  labs(y = "National Percentile", x = "Student Grade", title = "Literacy Skills")

```

```{r}
tested_students_by_school$ShortSchoolName <- sapply(strsplit(tested_students_by_school$SchoolName, " "), function(x) paste(x[1:min(length(x), 2)], collapse = " "))

tss <- ggplot(tested_students_by_school, aes(x = ShortSchoolName, y = num_students, fill = SchoolName)) +
  geom_col(color="black") +
  scale_fill_brewer(palette = "BrBu") +  my_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "italic")) +
  labs(x = "School Name", y = "Count of Students", 
       title = "Students Tested SY 2023")

tss

```

```{r}

tested_students_by_grade_school$ShortSchoolName <- sapply(strsplit(tested_students_by_grade_school$SchoolName, " "), function(x) paste(x[1:min(length(x), 2)], collapse = " "))



tsg <- ggplot(tested_students_by_grade, aes(x= StudentGrade, y = num_students, fill = StudentGrade)) +
  geom_col(color="black") +
  scale_fill_brewer(palette = "Dark2") +  my_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "italic")) +
  labs(x = "School Name", y = "Count of Students", 
       title = "Students Tested SY 2023")

tsg
```

```{r}
tested_students_
```
